---
title: "Rate analysis"
---

This article does a simple diversification rate analysis under the lineage-specific birth-death model. See first the installation vignette for a guide on how to install this program.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
First we load the necessary packages.
```{r setup, include=FALSE}
library(Pesto)
library(ape)
suppressPackageStartupMessages(library(ggtree))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidytree))
```
```{r setupfake, eval=FALSE}
library(Pesto)
library(ape)
library(ggtree)
library(ggplot2)
library(tidytree)
```
Next we load our favorite tree, and plot it.
```{r readtree}
data(primates)
th <- max(node.depth.edgelength(primates)) # the tree height
rho <- 0.67 # we know that the taxon sampling is around 2/3

ggtree(primates) +
  geom_tiplab(size = 2) +
  xlim(c(0.0, th + 10))
```

## Setting up the model

Our model will use three hyperparameters: $\hat{\lambda}$, $\hat{\mu}$ and $\eta$. What are appropriate values for $\hat{\lambda}$ and $\hat{\mu}$? One approach is to use so-called empirical hyperpriors for the parameters. Meaning, we will estimate $\hat{\lambda}$ and $\hat{\mu}$ using a simpler model: the constant-rate reconstructed birth-death process with uniform taxon sampling. 

```{r constantbdp, include=FALSE}
mlrates <- estimate_constant_bdp(primates, rho)

hatlambda <- mlrates$lambda
hatmu <- mlrates$mu
```
```{r constantbdpfake, eval=FALSE}
mlrates <- estimate_constant_bdp(primates, rho)

hatlambda <- mlrates$lambda
hatmu <- mlrates$mu
```

With a reasonable starting point for $\hat{\lambda}$ and $\hat{\mu}$, we can start setting up the state-dependent birth-death model. Suppose we want to allow $\lambda$ and $\mu$ to vary over a range of different values. Ideally, we would want to cover all possible values: the continuous range starting from 0.0 up to infinity. However, we do not now how to integrate over an infinitely large state space. Instead, we can select a finite sample of values. Let's say $\lambda$ and $\mu$ are log-normally distributed, and we pick $n$ points that are evenly spaced across the cumulative density function.

```{r lognormal}
n <- 6
H <- 0.578

lambda_quantiles <- lognorm_quantiles(meanlog = log(hatlambda), sdlog = H, n = n)
mu_quantiles <- lognorm_quantiles(meanlog = log(hatmu), sdlog = H, n = n)

## Let's pick all pairwise combinations among these quantiles
pairs <- allpairwise(lambda_quantiles, mu_quantiles)
lambda <- pairs[[1]]
mu <- pairs[[2]]
```

These values are now our $\vec{\lambda}$ and $\vec{\mu}$:
```{r plotlognormal, out.width = "80%"}
plot(lambda, mu, pch = 19, xlab = "lambda", ylab = "mu")
```

What do we do for $\eta$? One approach we can do is to find the maximum-likelihood estimate of $\eta$ conditional on $\vec{\lambda}$ and $\vec{\mu}$. The function `optimize_eta` will optimize for $\eta$ using gradient descent. This procedure typically needs to evaluate the likelihood (and the derivative of the likelihood with respect to $\eta$) function around 20-60 times before reaching convergence, so it is not instantaneous.

```{r estimate-eta}
eta <- optimize_eta(lambda, mu, primates, rho)
```

## Calculating the branch rates

Now that we have our parameters $\vec{\lambda}$, $\vec{\mu}$ and $\eta$, we can compute the expected rates across the branches:

```{r expected-rates}
td <- birth_death_shift(primates, lambda, mu, eta, rho)
```
The return object is a `treedata` object, which includes the tree and metadata. We can plot the results like so:
```{r treeplotrates, out.width = "80%"}
p1 <- ggtree(td, aes(color = `Speciation rate`)) +
             geom_tiplab(size = 2) +
             theme(legend.position = c(0.1, 0.8)) +
             xlim(c(0.0, th + 10))
plot(p1)
```

## Calculating the number of shifts

We can also calculate the expected number of shifts under the model.

```{r nshifts, out.width = "80%"}
td <- compute_nshifts(primates, lambda, mu, eta, rho)

p2 <- ggtree(td, aes(color = nshifts)) +
             geom_tiplab(size = 1) +
             theme(legend.position = c(0.1, 0.8)) +
             xlim(c(0.0, th + 10)) +
             scale_colour_gradient(low = "black", high = "red")
plot(p2)
```

Under the prior, we would expect $\eta \sum b_i$ number of shifts across the phylogeny, where $b_i$ is the branch length of branch $i$:
```{r priorN}
Nprior = eta * sum(primates$edge.length)
print(Nprior)
```
The posterior estimate of the number of state changes is highly correlated with the prior, but it is not exactly the same, it is slightly less: 
```{r posteriorN}
Nposterior = sum(td@data$nshifts)
print(Nposterior)
```
